---
title: ""
output: html_document
---

```{r setup, include=FALSE}

options(scipen=999)

library(dplyr)
library(lubridate)
library(tidyr)
library(reshape2)
library(RCurl)
library(rvest)
library(ggplot2)
library(scales)
library(astsa)
library(ggthemes)
library(extrafont)
library(reshape)
library(car)
library(forecast)
library(fGarch)
library(highcharter)
library(dygraphs)
library(xts)

#load data
emp <- read.csv(file = "C:/Users/GRA/Desktop/Misc/R Working Directory/Other/employment/formatted.emp.csv", strip.white = T)
emp$date <- as.character(emp$date)
emp$date <-as.POSIXct(strptime(emp$date, format = "%Y-%m-%d"))
emp$month <- month(emp$date, label = T)
emp.ts <- ts(emp$jobs, frequency = 12)
emp.change <- diff(emp.ts)
date2 <- emp$date[1:195]

sp <- read.csv(file = "C:/Users/GRA/Desktop/Misc/R Working Directory/Other/employment/formatted.sp.csv", strip.white = T)
sp$date <- as.character(sp$date)
sp$date <- ymd(sp$date)
sp$month <- month(sp$date, label = T)
sp.ts <- ts(sp$adj.close, frequency = 12)
sp.change <- diff(sp.ts)

manuf <- read.csv(file = "C:/Users/GRA/Desktop/Misc/R Working Directory/Other/employment/formatted.manuf.csv", 
                  strip.white = T)
manuf$date <- as.character(manuf$date)
manuf$date <- ymd(manuf$date)
manuf$month <- month(manuf$date, label = T)
manuf.ts <- ts(manuf$gacdfna, frequency = 12)
manuf.change <- diff(manuf.ts)

bai <- read.csv(file = "C:/Users/GRA/Desktop/Misc/R Working Directory/Other/employment/formatted.bai.csv", strip.white = T)
bai$date <- as.character(bai$date)
bai$date <- ymd(bai$date)
bai$month <- month(bai$date, label = T)
bai.ts <- ts(bai$nmfbai, frequency = 12)
bai.change <- diff(bai.ts)

cpi <- read.csv(file = "C:/Users/GRA/Desktop/Misc/R Working Directory/Other/employment/formatted.cpi.csv", strip.white = T)
cpi$date <- as.character(cpi$date)
cpi$date <- ymd(cpi$date)
cpi$month <- month(cpi$date, label = T)
cpi.ts <- ts(cpi$cpi, frequency = 12)
cpi.change <- diff(cpi.ts)

starts <- read.csv(file = "C:/Users/GRA/Desktop/Misc/R Working Directory/Other/employment/formatted.starts.csv", 
                   strip.white = T)
starts$date <- as.character(starts$date)
starts$date <- ymd(starts$date)
starts$month <- month(starts$date, label = T)
starts.ts <- ts(starts$starts, frequency = 12)
starts.change <- diff(starts.ts)


# highchart plotting paramaters
hc_params <- highchart() %>%
  hc_chart(zoomType = "x") %>%
  hc_tooltip(crosshairs = TRUE, shared = TRUE, borderWidth = 1, shared=T) %>%
  hc_yAxis(title=" ") %>%
  hc_exporting(enabled = TRUE) %>%
  hc_credits(enabled = TRUE,
             text = "BLS",
             href = "https://www.bls.gov/news.release/empsit.nr0.htm") %>% 
  hc_add_theme(hc_theme_gridlight()) %>%
  hc_legend(enabled = FALSE) %>%
  hc_subtitle(text="Since 2000", align="left") %>%
  hc_yAxis(title=" ")
alignment <- "left"  #subtitle alignment

```

# Forecasting the Monthly Change in Jobs {.tabset}
This page is my attempt to apply various time series methods to try and forecast US monthly seasonally adjusted, non-farm payrolls.  
The most recent jobs report from the BLS can be found [here.](http://www.bls.gov/news.release/empsit.nr0.htm)

## The Forecast 
Before developing a forecast I want to look at a few plots:  

* A  time series plot of the raw numbers
* A time series plot of the change in jobs (since that's what I'm trying to forecast)

The prominent decrease in payrolls from about 2008 to 2010 show a lagged effect of the great recession (officially December 2007 to June 2009). The number of non farm payrolls have been increasing since about October of 2010.  
```{r, echo=F}
emp.xts <- xts(emp$jobs, order.by=emp$date, frequency=12)
hc_params %>% 
  highchart(type = "chart",height = 500) %>% 
   hc_add_series_xts(emp.xts, name = "Seasonally adjusted non-farm payrolls") %>%
  hc_title(text = "Seasonally Adjusted Non-Farm Payrolls", align = alignment)
```

```{r, echo = F}
emp.change.date <- as.data.frame(emp$date)
emp.change.date <- as.data.frame(emp.change.date[-1,])

change.emp.xts <- xts(emp.change, order.by=emp.change.date$`emp.change.date[-1, ]`, frequency=12)

```

```{r, echo=F}
emp.acf <-

```


The change in jobs again shows the effect of the great recession. Since the great recession, the average change in non-farm payrolls has been `r format(mean(change.emp.xts[114:195,]*1000), digits=2)`.  

```{r, echo=F}
emp.change.date <- as.data.frame(emp$date)
emp.change.date <- as.data.frame(emp.change.date[-1,])

change.emp.xts <- xts(emp.change, order.by=emp.change.date$`emp.change.date[-1, ]`, frequency=12)
#mean(change.emp.xts)

(emp.change.plot <- hc_params %>% 
  highchart(type = "chart", height = 500) %>% 
  hc_add_series_xts(change.emp.xts, name = "Monthly Change in US Employment (000s") %>%
  hc_title(text = "Monthly Change in US Employment (000s)", align = alignment) %>%
  hc_subtitle(text = "mean change in monthly jobs = 66,179", align = alignment) %>%
  hc_yAxis(plotLines = list(
           list(label = list(text = ""),
                color = "#CD2626",
                width = 2,
                value = 66.1))))
```



## Covariates
description of covariates and plots

## Me1thodology +  Sources
methodology + sources
